Making Sequential Programs Deterministic

### 2019-04-06 design.tex

\todo{Move all of this to Section 3 since the core calculus has no type
  inference?
  But we do want to discuss \<sort>.  So maybe just move the discussion of type inference?}

\todo{The core calculus requires a qualifier on every type.}

\TheDeterminismChecker performs local type inference within method bodies.
It does not perform whole-program type inference.
This makes separate compilation possible.
It forces programmers to write specifications (type qualifiers) on methods,
which is good style and valuable documentation.

The local type inference is flow-sensitive.  That is, each variable may
have a different type on every line of the program, depending on
assignments to the variable.  The inferred type always respects any
declared type:  that is, it is always a refinement (a subtype) of the
declared type, if any.

In addition to assignments, some method calls refine the types of their
arguments.  One example is the \<sort> routine,
as shown in \cref{fig-sorting}.  One notable fact about the \rulename{sort}
rule is that it requires the type argument to be deterministic (type
qualifier \<Det>).  This is necessary because of two constraints.
First, refinement must not change the type argument, because of invariant
type argument subtyping.
Second, the only possible element type for a \<Det> collection is \<Det>.

\TheDeterminismChecker has additional type refinement rules for the sorting
methods of \<List>, \<Collections>, and \<Arrays> classes.
\begin{figure}
%     $\infer[\rulename{list sort}]{\Gamma \vdash x : \|Det \ List|\angles{\<Det> \ \beta}}{\Gamma \vdash x : \|OrderNonDet \ List|\angles{\<Det> \ \beta} & \|x.sort()|}$
%     
%     \bigskip

@@

    \caption{Sorting type refinement rules.  \<sort> is a method that works
      by side effect and changes the value---and the type---of its receiver.
    ``$x_{\mathrm{pre}}$ stands for a value before the call to \<sort>, and
      ``$x_{\mathrm{post}}$ stands for the value after the call.
    }
    \label{fig-sorting}
\end{figure}
As an example, if \codeid{sort()} is called on a receiver of type
\codeid{OrderNonDet List<Det Set<Det String>>}, it will be type refined to
\codeid{Det List<Det Set<Det String>>}. However, it is not permitted to
call \<sort> on a receiver of type \codeid{OrderNonDet List<OrderNonDet
  Set<Det String>>}.  The rule is not applicable, because doing so would
instantiate invalid types.

>>>>

\todo{I feel that this section is very confusing.}

Some method calls refine the types of their
arguments.  One example is the \<sort> routine,
as shown in \cref{fig-sorting}.  One notable fact about the \rulename{sort}
rule is that it requires the type argument to be deterministic (type
qualifier \<Det>).  This is necessary because of two constraints.
First, refinement must not change the type argument, because of invariant
type argument subtyping.
Second, the only possible element type for a \<Det> collection is \<Det>.

\begin{figure}
%     $\infer[\rulename{list sort}]{\Gamma \vdash x : \|Det \ List|\angles{\<Det> \ \beta}}{\Gamma \vdash x : \|OrderNonDet \ List|\angles{\<Det> \ \beta} & \|x.sort()|}$
%     
%     \bigskip

@@

    \caption{Sorting type refinement rules.  \<sort> is a method that works
      by side effect and changes the value---and the type---of its receiver.
    ``$x_{\mathrm{pre}}$ stands for a value before the call to \<sort>, and
      ``$x_{\mathrm{post}}$ stands for the value after the call.
    }
    \label{fig-sorting}
\end{figure}

For example, if \codeid{sort()} is called on a receiver of type
\ptype{OrderNonDet List}{\lptype{Det Set}{Det String}}, it will be type refined to
\ptype{Det List}{\lptype{Det Set}{Det String}}. However, it is not permitted to
call \<sort> on a receiver of type \ptype{OrderNonDet List}{\lptype{OrderNonDet
  Set}{Det String}}.  The rule is not applicable, because doing so would
instantiate invalid types.

### 2019-04-06 java-design.tex

>>>>

\todo{
\OurTypeSystem performs local type inference within method bodies.
It does not perform whole-program type inference.
This makes separate compilation possible.
It forces programmers to write specifications (type qualifiers) on methods,
which is good style and valuable documentation.

The local type inference is flow-sensitive.  That is, each variable may
have a different type on every line of the program, depending on
assignments to the variable.  The inferred type always respects any
declared type:  that is, it is always a refinement (a subtype) of the
declared type, if any.
}
